Discrepancy, Coresets, and Sketches in Machine Learning

### 2019-01-15 komlos.tex
appendinx MRL

\begin{abstract}
This paper defines the discrepancy for families of functions.  It shows that low discrepancy classes admit small offline and streaming coresets.
We provide general techniques for bounding the discrepancy of machine learning problems and specifically does so for matrix covariance approximation, logistic regression, kernel density and any analytic function of the dot product or the squared distance.
Our result resolves a long standing open question regarding minimal cardinality coresets for Gaussian kernel density estimation.  
We provide two more related but independent results. 
First, an exponential improvement of the widely used merge-and-reduce trick which gives improved streaming sketches for any low discrepancy problem.
Second, an extremely simple algorithm for finding coresets (low discrepancy sequences) for any positive kernel density estimation problem. 
This paper makes an explicit connection between low discrepancy, coresets, learnability, and streaming algorithms. 
\end{abstract}

\section{Introduction}
In both machine learning and in streaming and sketching problems our goal is (often) to approximate sums or expectations of well behaved functions.
Specifically, we need to approximate $\E_x f(x)$ or $\frac{1}{n}\sum_{i=1}^{n} f(x_i)$ for every $f\in \F$ where $\F$ is a family of functions and $x_i \in \mathcal X$ are either sampled training examples or an arbitrary set of stream items. 
Standard generalization results show that for a large enough value of $n$ the average approximates the mean if the complexity of $\F$ is bounded and the samples $x_i$ are i.i.d.\ drawn from underlying distribution. We therefore focus on approximating the average, or rather, the sum $\sum_{i=1}^{n} f(x_i)$. For notational convenience we use a parameter $q \in \mathcal Q$ to index into $\F$ explicitly $F(q) = \sum_{i=1}^{n} f(x_i, q)$ . One should think about $q$ as either the model parameters or a query for the sketch.

@@

\begin{theorem} \label{thm:streaming12}
For any function family $\F$ with a corresponding discrepancy $K_m = O(c/m)$ there exists an fully-mergeable streaming coreset deterministic algorithm of size $O\left(c\log^2(\eps n/c)/\eps\right)$ whose error is at most $\eps n$.
\end{theorem}
\begin{theorem} \label{thm:streaming11}
For any function family $\F$ with a corresponding discrepancy $K_m = O(c/m)$ there exists an fully-mergeable streaming coreset randomized algorithm of size $O\left(c\log^2\log(n/\delta)/\eps\right)$ whose error is at most $\eps n$ with probability at least $1-\delta$. \el{for a single query?}
\end{theorem}

@@

\section{Proofs for Section~\ref{sec:sketch}, Sketching Coresets}

\begin{lemma} \label{lem:compactor}
Assume we have black box access to a solver that for any integer $m$ and inputs $x_1,\ldots,x_m$ obtains signs $\sigma_1,\ldots,\sigma_m$ such that $\max_q \left| \sum_{i=1}^{m} \sigma_i f(x_i, q)\right| \leq K_m$ for some function $f$ whose discrepancy is $K_m$.\footnote{An constant factor approximation algorithm achieving $\max_q \left| \sum_{i=1}^{m} \sigma_i f(x_i, q)\right| = O(K_m)$ would also clearly suffice.} 
Then there exist a streaming algorithm requiring a memory buffer of $m$ input items that given a stream of length $n$ outputs a stream $z_1,\ldots,z_{n'}$ with the following properties
\begin{itemize}
\item $\{z_i\}_i$ is a subset of $\{x_i\}$
\item $\E[n'] = n/2$
\item For $\rho >0$, $\Pr[n' \geq (1+\rho)n/2] \leq \exp \left( -O\left(\frac{n\rho^2}{m}\right)\right)$
\item For any fixed query $q$, the error associated with $q$, defined as
$$\text{Err}(q) = 2\sum_{i=1}^{n'} f(z_i,q) - \sum_{i=1}^n f(x_i,q)  $$
Can be decomposed as a sum
$$\text{Err}(q) = \sum_{j=1}^{n/m} \text{Err}_j(q)$$
where the different $\text{Err}_j(q)$ are independent, have mean 0 and $|\text{Err}_j(q)| \leq K_k$
\end{itemize}
\end{lemma}

>>>>

\begin{abstract}
This paper defines the discrepancy for families of functions.  It shows that low discrepancy classes admit small offline and streaming coresets.
We provide general techniques for bounding the discrepancy of machine learning problems and specifically does so for matrix covariance approximation, logistic regression, kernel density and any analytic function of the dot product or the squared distance.
Our result resolves a long standing open question regarding minimal cardinality coresets for Gaussian kernel density estimation.  
We provide two more related but independent results. 
First, an exponential improvement of the widely used merge-and-reduce trick which gives improved streaming sketches for any low discrepancy problem.
Second, an extremely simple algorithm for finding low discrepancy sequences (and therefore coresets) for any positive semi-definite kernel. 
This paper makes an explicit connection between low discrepancy, coresets, learnability, and streaming algorithms. 
\end{abstract}

\section{Introduction}
In both machine learning and in streaming and sketching problems our goal is (often) to approximate sums or expectations of well behaved functions.
Specifically, we need to approximate $\E_x f(x)$ or $\frac{1}{n}\sum_{i=1}^{n} f(x_i)$ for every $f\in \F$ where $\F$ is a family of functions and $x_i \in \mathcal X$ are either sampled training examples or an arbitrary set of stream items. 
Standard generalization results show that for a large enough value of $n$ the average approximates the mean if the complexity of $\F$ is bounded and the samples $x_i$ are drawn i.i.d.\ from underlying distribution. We therefore focus on approximating the average, or rather, the sum $\sum_{i=1}^{n} f(x_i)$. For notational convenience we use a parameter $q \in \mathcal Q$ to index into $\F$ explicitly $F(q) = \sum_{i=1}^{n} f(x_i, q)$. 
One should think about $q$ as either the model parameters or a query for the sketch.

@@

\begin{theorem} \label{thm:streaming11}
For any function family $\F$ with a corresponding discrepancy $K_m = O(c/m)$ there exists an fully-mergeable streaming coreset deterministic algorithm of size $O\left(c\log^2(\eps n/c)/\eps\right)$ whose error is at most $\eps n$.
\end{theorem}
\begin{theorem} \label{thm:streaming12}
For any function family $\F$ with a corresponding discrepancy $K_m = O(c/m)$ there exists an fully-mergeable streaming coreset randomized algorithm of size $O\left(c\log^2\log(n/\delta)/\eps\right)$ whose error is at most $\eps n$ with probability at least $1-\delta$. \el{for a single query?}
\end{theorem}

@@

\section{Appendix Appendix Appendix...}

The proofs of Theorems \ref{thm:streaming11}, \ref{thm:streaming12}, \ref{thm:streaming21}, and \ref{thm:streaming22} all use the basic concept of a compactor. A compactor consumes a stream of items and outputs another stream. 
The output stream contains roughly half the items from the input stream with double the weight. 
It does so by keeping a buffer of a certain capacity $k$. When a new item is inserted into the compactor it is added to its buffer. 
If the buffer is full, a compaction operation takes place. 
The compaction takes the elements in the buffer $x_1,\ldots,x_m$ and finds a low discrepancy assignment $\sigma$ such that 
$\max_q |\sum_i \sigma_i f(x_i,q)| = O(mK_m)$. 
Note that such a sequence is guarantied to exist by the definition of the discrepancy. 
However, an algorithm for finding such a sequence might not exist. 
From this point onwards, we assume the existence of such an algorithm.
In other words, our theorems hold for problems which admit such a solution.
The compactor then outputs either $\{ x_i | \sigma_i = 1\}$ or  $\{ x_i | \sigma_i = -1\}$ with double the weight.

We begin with showing some basic properties of deterministic compactor, which always output the smaller of the two set  $\{ x_i | \sigma_i = 1\}$ or  $\{ x_i | \sigma_i = -1\}$. Consider a single compaction and let $\tilde F$ denote the function evaluated on the output set.
Let $E(q) = \sum_i \sigma_i f(x_i,q)$ for the signs $\sigma$ computed by the algorithm above. 
$$\tilde F(q) = \sum_{i ,\; \sigma_i=1} 2f(x_i, q) = \sum_{i} f(x_i, q) +  \sum_{i} \sigma_i f(x_i, q) = F(q) + E(q)$$
$$\tilde F(q) = \sum_{i ,\; \sigma_i=-1} 2f(x_i, q) = \sum_{i} f(x_i, q) - \sum_{i} \sigma_i f(x_i, q) = F(q) - E(q)$$
Note that $|\tilde F(q) - F(q)| = |E(q)| \le \max_q |\sum_i \sigma_i f(x_i,q)| = O(mK_m)$. 
Now, aggregating over the $n/m$ compactions we get that $|\tilde F(q) - F(q)| \le n K_m$. Moreover, the stream length is clearly cut (at least) in half.

These facts alone already allow us to prove Theorem~\ref{thm:streaming11}. 
\paragraph{Theorem~\ref{thm:streaming11}}
For any function family $\F$ with a corresponding discrepancy $K_m = O(c/m)$ there exists an fully-mergeable streaming coreset deterministic algorithm of size $O\left(c\log^2(\eps n/c)/\eps\right)$ whose error is at most $\eps n$.
\begin{proof}
Consider feeding the output of the first compactor into a second one etc. Number the compactors $0,\ldots,H$.
The weight of item to compactors $h$ have weight $w_h = 2^h$. The length of the input stream seen by compactor is $n_h \le n/2^h$.
Each compactor contributes at most $w_h n_h K_m \le n K_m$ error. Moreover since $n_h \le n/2^h$ we know that $H \le log(n/m)$.
The total error is therefore $H n K_m \le log(n/m) n K_m$. 
Setting $m \ge c\log(\eps n/c)/\eps$ and replacing $K_m = c/m$ we get that the error is at most $log(n/m) n K_m \le \eps n$ as required. 
Since we have $H \le log(\eps n)$ such compactors the overall space complexity is $c\log^2(\eps n/c)/\eps$.
\end{proof}

This algorithm is a direct extension the well know MRL algorithm \cite{MRL} for quantile sketching. 
Note that for quantiles, $f(x,q) = 1$ if $q > x$ and $0$ else. 
A low discrepancy sequence is achieved simply by sorting the values and assigning $\sigma_i = 1$ for all evenly positioned values in the sorted order and $\sigma_i=-1$ to the odd positions. The above gives a discrepancy of $1/m$ for quantile approximation.

\section{Proofs for Section~\ref{sec:sketch}, Sketching Coresets} 
\begin{lemma} \label{lem:compactor}
Assume we have black box access to a solver that for any integer $m$ and inputs $x_1,\ldots,x_m$ obtains signs $\sigma_1,\ldots,\sigma_m$ such that $\max_q \left| \sum_{i=1}^{m} \sigma_i f(x_i, q)\right| \leq K_m$ for some function $f$ whose discrepancy is $K_m$.\footnote{An constant factor approximation algorithm achieving $\max_q \left| \sum_{i=1}^{m} \sigma_i f(x_i, q)\right| = O(K_m)$ would also clearly suffice.} 
Then there exist a streaming algorithm requiring a memory buffer of $m$ input items that given a stream of length $n$ outputs a stream $z_1,\ldots,z_{n'}$ with the following properties
\begin{itemize}
\item $\{z_i\}_i$ is a subset of $\{x_i\}$
\item $\E[n'] = n/2$
\item For $\rho >0$, $\Pr[n' \geq (1+\rho)n/2] \leq \exp \left( -O\left(\frac{n\rho^2}{m}\right)\right)$
\item For any fixed query $q$, the error associated with $q$, defined as
$$\text{Err}(q) = 2\sum_{i=1}^{n'} f(z_i,q) - \sum_{i=1}^n f(x_i,q)  $$
Can be decomposed as a sum
$$\text{Err}(q) = \sum_{j=1}^{n/m} \text{Err}_j(q)$$
where the different $\text{Err}_j(q)$ are independent, have mean 0 and $|\text{Err}_j(q)| \leq K_k$
\end{itemize}
\end{lemma}
